Inference of ancestral locations

### 2024-06-22 manuscript/phylogeo.tex
small edits

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

@@

\begin{figure*}[tb]
    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
    % \includegraphics*[width=0.48\textwidth]{figures/yule_dispersal.pdf}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities. Their numerical value depends strongly on sample size and there is no ground truth to compare to. The expected scaling $\sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
\end{figure*}

@@

Instead of the coalescent trees used above, I now used an individual based simulation forward in time where individuals are assigned a fitness $f=\alpha(1-\hat{\phi}/\phi_0) + (1-N/N_0)$ and produce a Poisson distributed number of offspring with mean $1+f$.
The first term determines the fitness due to the local population density and has the tendency to stabilize this density at $\phi_0$, the second term $(1-N/N_0)$ keeps the overall population close to the target population size $N_0$.
The population density is calculated using a Gaussian kernel with interaction radius $r$
\begin{equation}
    \ref{eq:density_estimator}
    \hat{\phi}(\rvec, t) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{\sqrt{2\pi r^2}}
\end{equation}
where the sum is over all individuals $i$ in the population of size $N$ and $\rvec_i$ is location of individual $i$ at time $t$.
The spatial locations of the offspring are chosen from a 2-D Gaussian centered on the parent with a variance to match the diffusion constant.
The simulation records the tree of population, including their ancestral location and times.
In addition to the true location of ancestral nodes, I also infer their locations from the locations of the extant samples.
Since the model is purely diffusive, the marginal distributions of ancestral locations and their mean and variance can be calculated exactly using recursion relations.

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
\usepackage{hyperref}
%\usepackage{cite}

@@

\begin{figure*}[tb]
    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
    % \includegraphics*[width=0.48\textwidth]{figures/yule_dispersal.pdf}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
    While the diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities. For Kingman coalescent tree their numerical value depends strongly on sample size and there is no ground truth to compare to. The expected scaling $v_w \sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
\end{figure*}

@@

Instead of the coalescent trees used above, I now used an individual based simulation forward in time where individuals are assigned a fitness $f=\alpha(1-\hat{\phi}/\phi_0) + (1-N/N_0)$ and produce a Poisson distributed number of offspring with mean $1+f$.
The first term determines the fitness due to the local population density and has the tendency to stabilize this density at $\phi_0$, the second term $(1-N/N_0)$ keeps the overall population close to the target population size $N_0$.
The population density is calculated using a Gaussian kernel with interaction radius $r$
\begin{equation}
    \label{eq:density_estimator}
    \hat{\phi}(\rvec, t) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{\sqrt{2\pi r^2}}
\end{equation}
where the sum is over all individuals $i$ in the population of size $N$ and $\rvec_i$ is location of individual $i$ at time $t$.
The spatial locations of the offspring are chosen from a 2-D Gaussian centered on the parent with a variance to match the diffusion constant.
The simulation records the tree of population, including their ancestral location and times.
In addition to the true location of ancestral nodes, I also infer their locations from the locations of the extant samples.
For this inference, I make the common assumption of continuous phylogeography that growth and location are uncoupled and that dispersal is diffusive.
In this case, the marginal distributions of ancestral locations and their mean and variance can be calculated exactly using recursion relations.

@@

\section*{Acknowledgements}
I am grateful to Louis de Plessis and Oskar Hallatscheck for stimulating discussions and useful feedback on earlier versions of this manuscript.

### 2024-06-22 notes/most_likely_diffusion.tex
small edits

>>>>

\section{common examples}

\subsection{WNV}
West-Nile virus was first detected on the east coast of the US in 1999 and reached the West Coast 5 years later in 2004. This translates to a wave front velocity of about 1000 km/year. Previous analysis estimated a diffusion coefficient of between 200 and 10000 km^2/day (Pybus et al). Using $v_{FKPP} = \sqrt{2D\alpha}$ and equating it to 1000 km/year = 3km/day, we obtain ranges of $\alpha=\frac{10 km^2}{2D day^2} = 0.025/day \ldots 0.0005/day$ or between $0.0035$ and $0.14$/week.
A pathogen that produces seasonal outbreaks with many orders of magnitude in prevalence between peak and troughs typically requires a growth rate of 1/week, considerably more than what the estimates of $D$ and the FKPP equation imply.
Alternatively, if we assume $\alpha = 0.1/day$, we would expect $D = \frac{10 km^2}{0.2 day} = 50 km^2/day$, implying a daily exploration radius of about 10km.
This picture is also consistent with the apparent ``slowing down" if lineages after the initial expansion across North America: Once the habitat was fully explored, directed invasion with $v_{FKPP}$ ceases.

\subsection{rabies}
Dispersal rates of rabies are typically estimated to be between 500 and 1500 km^2/year.
The most recent common ancestor of various populations is between 50 and 100 years in the past.
This would translate into rather limited spread of 200 to 500km of individual clades from their common ancestor, suggesting that these populations are highly fragmented and their long range dispersal is dominated by rare introductions of the virus into new populations.
Within each population, phylogeographic reconstruction will often give plausible results, but it is hard to extrapolate to longer time and distance scales.
